#Loan analysis by Alona Sukretna
========================================================
```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that using in analysis in this code chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

# Plotting
library(ggplot2)
# Scaling for plots 
library(scales)
# Filtering and grouping
library(dplyr)
# Scatterplot matrices
library(GGally)
# Date manipulation
library(lubridate)
# Tabulate modules
library(memisc)
#Formatting plots
library(grid)
library(gridExtra)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
mydata <- read.csv(file="prosperLoanData.csv", header=TRUE)
```
**Data set:** I choose Loan Data from Prosper 
(https://s3.amazonaws.com/udacity-hosted-downloads/ud651/prosperLoanData.csv).
This data set contains 113,937 loans with 81 variables on each loan,
including loan amount, borrower rate (or interest rate), current loan status,
borrower income, borrower employment status, borrower credit history, 
and the latest payment information.

# Univariate Plots Section
**Preliminary exploration of dataset**: There are a lot of NA data in our dataset,
but there is reasons why data is empty. For example, different stage of loans, 
applied methodics and so on. So, we'll be doing data cleaning if needed during 
our exploration.

First, let's explore main parameteres for loans, e.g. amount, term, current status.

###1. Loan Original Amount
```{r echo=FALSE, LoanOriginalAmount}
summary(mydata$LoanOriginalAmount)
```
We have loans in our data set that starts from $1000 up to $35,000.
The average loan amount is $8,337.

```{r echo=FALSE, LoanOriginalAmountPlot}
ggplot(data=mydata, aes(LoanOriginalAmount))+
         geom_histogram(binwidth = 700)+
         scale_x_continuous(breaks=seq(0, 26000,2000), limits=c(0,26000))
```

```{r echo=FALSE, LoanAmountSummary}
 summary(mydata$LoanOriginalAmount)
```

We can see that the largest count of loans for $4,000. Also we have a few peak 
counts for $2,000, $3,000, $5,000, $10,000, $15,000, $20,000, $25,000.
Probably, the greateast count of $4000 loans is due to different procedures and 
requirements for larger amounts.
There is a few loans larger $25,000 but for more clear view we can exclude them
from our plot.
The mean loan amount is $8337.

###2.Loan Term

```{r echo=FALSE, LoanTermPlot}
#converting to factor data
mydata$Term<-factor(mydata$Term)

ggplot(data=mydata, aes(Term))+
         geom_bar()
```

Most loans are for 36 months(3 years), less for 60 months(5 years),
and the least for 12 moths.

###3.Loan Current Status

```{r echo=FALSE, LoanCurrentStatusPlot}
ggplot(data=mydata, aes(LoanStatus))+
          geom_bar()+
          theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

Obviously, the most loans are in Current status. There are a lot of Completed 
loans in our data set. A big count of loans are Chargeoff (no payments for 6 
month, unlikely to be collected) and Defaulted. Past Due loans have own gradation.
Tere is small count of such loans but I prefer not exclude them from analysis 
because they connected with further analyzed parameters.

```{r echo=FALSE, LoanCurrentStatusTable}
table(mydata$LoanStatus)
```

###4. Aim of the loan(Listing Category)

The category of the listing that the borrower selected when posting their listing:
0 - Not Available, 1 - Debt Consolidation, 2 - Home Improvement, 3 - Business,
4 - Personal Loan, 5 - Student Use, 6 - Auto, 7- Other, 8 - Baby&Adoption, 
9 - Boat, 10 - Cosmetic Procedure, 11 - Engagement Ring, 12 - Green Loans,
13 - Household Expenses, 14 - Large Purchases, 15 - Medical/Dental,
16 - Motorcycle, 17 - RV, 18 - Taxes, 19 - Vacation, 20 - Wedding Loans.

```{r echo=FALSE, Aim_of_the_loan(ListingCategory)}
ggplot(data=mydata, aes(ListingCategory..numeric.))+
  geom_bar(show.legend=TRUE)+
  scale_x_continuous(breaks = c(0,1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20))
```

The most loans are for Debt Consolidation. For big count of loans information
not available or not in listed categories ('others').

###5.Loan Origination Date
It's in format "3/3/2014  12:00:00 AM", we don't need time for analysis.

```{r echo=FALSE, new_variable_for_Loan_Origination_Date}
#create new variable for date without time
mydata$LoanDate<-substr(as.character(mydata$LoanOriginationDate),1,7)
#date format
mydata$LoanDate<-as.Date(paste(mydata$LoanDate,"-01", sep=""))
#plot
ggplot(mydata, aes(LoanDate))+
  geom_bar()
```

We can see that we have data from the end of 2005 to the 2014.
Company loans grows for this period with a few ups and downs and 1 gap around 
2009. This gap can be explained that company has changed it's business model
at that time(2008 'quiet period' and 2009 SEC registration according to Wikipedia)

###6. ProsperRating..numeric. 

```{r echo=FALSE, LoanProsperRating..numericPlot}
ggplot(data=mydata, aes(ProsperRating..numeric.))+
         geom_bar()+
         scale_x_discrete(limits=c("HR","E","D","C","B","A","AA"))
```

We have most loans with average ProsperRating and less with the highest and lowest
ratings.

###7. CreditScoreRangeLower

We have two indicators of Credit Score:CreditScoreRangeLower and CreditScoreRangeUpper.
I plot it together to see the difference.

```{r echo=FALSE, CreditScoreRangeLowerPlot}
ggplot(data=mydata, aes(x=CreditScoreRangeLower, fill = "red",
                                    alpha = 0.2))+
         geom_histogram()+ 
         scale_x_continuous(breaks=seq(400,950,100), limits=c(400,950)) +
        geom_histogram(data=mydata, aes(x=CreditScoreRangeUpper, fill = "blue",
                                        alpha = 0.2))+
   scale_fill_discrete(name="Credit Scores", labels=c("Lower", "Upper"))+
   labs(x = "Credit Scores")
   
 
```
```{r echo=FALSE, CreditScoreRangeLowerBoxPlot}
grid.arrange(ggplot(aes(x=1, y=CreditScoreRangeLower), 
        data = mydata) +
   geom_boxplot( ), 
   ggplot(aes(x=1, y=CreditScoreRangeUpper), 
        data = mydata) +
   geom_boxplot( ), nrow=1)
  
```

I change scale because there was an outlier around 0. And there is a few outliers
more than 750 and less than 640.

For further analysis we'll use CreditScoreRangeLower, because it's a minimum 
acceptable value that mentioned in Prosper rules.
Or alternatively we can create additional value that is average of lower and 
upper scores.

The plot shows that in average credit scores veries from 640 to 750 and further
in most loans. This align with rule that was implemewnted by Prosper in 2009 that
minimum Credit Score required to be 640. This also explain small amount of loans
with credit score lower 600, that for sure was isuued before 2009 and lead to 
revision of rules.


###8. Estimated Loss
Estimated loss is the estimated principal loss on charge-offs. 
This parameter is applicable for loans originated after July 2009.

```{r echo=FALSE, EstimatedLossPlot}
grid.arrange(ggplot(mydata, aes(EstimatedLoss))+
  geom_histogram()+
  scale_x_continuous(breaks=pretty_breaks(n=10), limits=c(0,0.2)),
ggplot(aes(x=1, y=EstimatedLoss), 
        data = mydata) +
  geom_boxplot( ), nrow =1)
         
```
 
Based on numbers I think it's not EstimatedLoss amount, but EstimatedLoss Rate 
(which is not obvious from Variable Definition that comes with data).
We have a few loans with EstimatedLoss larger than 0.3, so we can eliminate them 
to see other data more clear. Also I wanted to see which loans has bigger
EstimatedLoss. According to our plot and summary statistics(below) we can see 
that the biggest number of loans have EstimatedLoss around 0.06.

```{r echo=FALSE, EstimatedLossSummary}
 summary(mydata$EstimatedLoss)
```

Based on summary statistics we can see that the average loss is 0.08. Maximum loss
is 0.336, but this data is outlier, so it's rather exception.

###9. Borrower Rate

```{r echo=FALSE, BorrowerRatePlot}
grid.arrange( ggplot(mydata, aes(BorrowerRate))+
  geom_histogram()+
  scale_x_continuous(breaks=pretty_breaks(n=10), limits=c(0,0.4)),
ggplot(aes(x=1, y=BorrowerRate), 
        data = mydata) +
  geom_boxplot( ), nrow =1)
         
```

```{r echo=FALSE, BorrowerRateSummary}
 summary(mydata$BorrowerRate)
```

Based on plots and summary we can see that average borrower rate is 0.19. Peak at
rate 0.31-0.32. I think is for not reliable loans with bad scores and around 0.15
for reliable loans with good scores.
There is a few outliers with Borrower Rate more than 0.4. It's rather exceptions,
so we can exclude them from analysis.


###10. Borrower APR

```{r echo=FALSE, BorrowerAPRPlot}
grid.arrange( ggplot(aes(x=BorrowerAPR), 
        data = mydata) +
  geom_histogram( bins = 30) ,
   ggplot(aes(x=1, y=BorrowerAPR), 
        data = mydata) +
  geom_boxplot( )  , nrow =1)
         
```
```{r echo=FALSE, Borrower_Rate_Summary}
 summary(mydata$BorrowerRate)
```

Borrower APR( Borrower's Annual Percentage Rate) is in range between 0 and 0.4975.
The most popular APR is around 0.2, mean is 0.1928. There is a few outliers 
in APR more than 0.4, and we can exclude them from further analysis.

# Univariate Analysis

 **Summarizing data**: 

We have big data set with loan data from Prosper from the end of  2005 to 2014. 
It's quite representative data set with lot of variables.
We don't need it all for analysis, but better to have more, than less. 
I'll make needed changes in the dataset during my investigtion. For example,
I created new variable for loan date without time for analysing loan history.

We are interested in loan distribution by year, quality, usage to see all changes.
We need to find what are the factors that influence the most on loan quantity
and quality.

Also I analyzed a few parameters that I'll use in further analysis: Prosper Rating,
Credit Score, Estimated Loss and Borrower Rate.

Besides analyzed paramateres, also would be interesting to see geography  of loan
distribution (by BorrowerState), characteristic of people using Prosper
(occupation, income), monthly payments.



# Bivariate Plots Section

> **Choosing variables might be interesting to look**: 
Based on our data set and what we saw in the univariate plots, let's choose some 
variables for bivariate analysis. 
First, we will look on correlation between parameters. The matrix is so big that
I preffer not to show it in html.

```{r echo=FALSE, Bivariate_Plots_correlation, results="hide"}
#numeric columns
#sapply(mydata, is.numeric)
#use boolean vector to subset and get just the numeric columns
numeric_columns<-mydata[, sapply(mydata, is.numeric)]
#correlation matrix
res <- cor(numeric_columns, use = "complete.obs", method = "pearson")
round(res, 2)
#getOption("max.print") limited output of correlation matrix in html, it's big
#options(max.print=500)
```

We can see significant correlation(positive or negative) between BorrowerRate,
LenderYield, EstimatedLoss, ProsperRating..numeric, Prosper Score,
Credit Score(Lower and Upper) and APR. 
This is logical because different credit ratings and scores should be comparable
to each other. And the worst rating is the bigger will be estimated loss. 
Also obviously big enough correlation between Loan Amount and Monthly Payment.
This matrix is too big. Let's see this parameters more closely.

###1. Prosper Rating vs Credit Score.

```{r echo=FALSE, Bivariate_Plots_cortest}
cor.test(mydata$ProsperRating..numeric.,mydata$CreditScoreRangeLower)

```

According to Wikipedia:"Since its SEC registration in 2009, Prosper has provided
a proprietary "**Prosper Rating**" for prospective borrowers based on the
company's estimation of that borrower's "estimated loss rate." According to the 
company, that figure is "determined by two scores: (1) the **credit score**,
obtained from an official credit reporting agency, and (2) the **Prosper Score**, 
figured in-house based on the Prosper population." Prosper Ratings, from
lowest-risk to highest-risk, are labeled AA, A, B, C, D, E, and HR ("High Risk")."

```{r echo=FALSE, Bivariate_Plots1}
p1<-boxplot(CreditScoreRangeLower~factor(ProsperRating..numeric.), data=mydata, 
        col = c("pink", "blue"),  xlab = "ProsperRating",  ylab = "CreditScoreRangeLower")
p1       
```

This boxplot shows the relationship between borrower's Prosper rating
(after July 2009) and CreditScore(Lower). CreditScore is a part of Prosper Rating 
along with Prosper Score. The higher Rating (7 is the highest) the higher 
should be Credit Score. Just for rating 1 and 2 we have some exclusions.


```{r echo=FALSE, Bivariate_Plots_summary}
#by(mydata$CreditScoreRangeLower,mydata$ProsperRating..numeric., summary)
```

Notes:
**CreditScoreRangeLower**	The lower value representing the range of the 
borrower's credit score as provided by a consumer credit rating agency.
CreditScoreRangeUpper	The upper value representing the range of the borrower's
credit score as provided by a consumer credit rating agency. 
I used lower level for calculations but we also can create mean CreditScore from
Lower and Upper.

**ProsperRating (numeric)**	The  Prosper Rating assigned at the time the listing
was created: 0 - N/A, 1 - HR, 2 - E, 3 - D, 4 - C, 5 - B, 6 - A, 7 - AA.
Applicable for loans originated **after July 2009**.
ProsperRating (Alpha)	The Prosper Rating assigned at the time the listing was
created between AA - HR.  Applicable for loans originated after July 2009.

Our findings align with information from wikipedia, that 
" New prospective borrowers were required to have
an FICO 8 credit score of at least 640."

###2. Prosper rating	vs Estimated Loss

```{r echo=FALSE, Bivariate_Plots_cor}
cor.test(mydata$ProsperRating..numeric.,mydata$EstimatedLoss)

```

Negative correlation,  one variable increases as the other decreases.

```{r echo=FALSE, Bivariate_Plots_2_boxplot}
p2<-boxplot(EstimatedLoss~factor(ProsperRating..numeric.), data=mydata, 
        col = c("pink", "blue"),  ylim = c(0, 0.2), xlab = "ProsperRating",
         ylab = "EstimatedLoss")
p2
```

```{r echo=FALSE, Bivariate_Plots_summary2}
#by(mydata$EstimatedLoss,mydata$ProsperRating..numeric., summary)
```

For the worst Rating (1) EstimatedLoss will be the most sufficient.

###3. Prosper Rating vs Loan Amount

```{r echo=FALSE, Bivariate_Plots_cor_test}
cor.test(mydata$ProsperRating..numeric.,mydata$LoanOriginalAmount)

```

Correlation is not so strong.

```{r echo=FALSE, Bivariate_Plots3}
ggplot(data = mydata, aes(x=ProsperRating..numeric., y=LoanOriginalAmount))+
         geom_jitter(aes(color=ProsperRating..numeric.), width = 0.3, height = 0.3)+
         scale_x_discrete(limits=c("HR","E","D","C","B","A","AA"))+
         scale_color_continuous(high='red',low='blue')

```

This plot is overplotted so I use a boxplot for the same parameters.

```{r echo=FALSE, Bivariate_Plots_2}
p3<-boxplot(LoanOriginalAmount~factor(ProsperRating..numeric.), data=mydata, 
        col = c("pink", "blue"), xlab = "ProsperRating",
         ylab = "Loan Amount",
         ylim = c(0, 25000) )
p3
```

I set some limits for the big amount ouliers because it's rather exceptions.
Loans with high ammount of money has higher ratings, that's mean less risky.
High risk loans (HR Rating - 1) are loans for small amounts of money(up tp $10,000).





# Bivariate Analysis

In this section I analize how Prosper Rating (this characteristic was created by Prosper)
connected with Credit Score, EstimatedLoss and Loan Amount.
The strongest negative correlation is between Prosper Rating
and Estimated Loss. It's obvious that the better raiting the less loss. Also it's 
closely related with Credit Score, which is part of Prosper Rating calculating.
And there is enough dependency between Prosper Rating and Loan Amount, because bigger 
loans give only for reliable clients. Also this rating should be strongly connected with 
Borrower Rate and Borrower APR.
There is also strong relationship between term and loan amount: longer term,
bigger amount.



# Multivariate Plots Section

In previous section we found thar Prosper Rating connected with
Credit Score, Estimated Loss, Loan Amount and Borrower Rate, APR.
Let's continue our investigations.

```{r echo=FALSE, Multivariate_Plots1}
#subset NA values for ProsperRAting
#we have missing data for Cancelled, ChargedOff, Completed and 
#Defaulted loans. Probably, for such loans rating not calculated.
#Let's use cleaned data:

mydata1<-subset(mydata, !is.na(ProsperRating..numeric.))

#ggplot(aes(x=EstimatedLoss, y=CreditScoreRangeLower, 
# color=factor(ProsperRating..numeric.)), data=mydata1)+
#geom_jitter()+ylim(550,900)+scale_color_brewer(type = 'seq', palette = 'YlGnBu')+
 #theme_gray(base_size = 14)

  
ggplot(mydata1, aes(x=EstimatedLoss, y=CreditScoreRangeLower, 
                    fill=factor(ProsperRating..numeric.), colour=factor(ProsperRating..numeric.))) +
#  geom_point() +
  geom_boxplot(alpha=0.6)

```

Here we can see again that low Prosper rating(1) leads to bigger Estimatedloss while 
CreditScore sometimes can be high but have  bad Prosper Rating and
high loss. This is one of the reasons, I think, that Prosper use their own rating,
in which Credit Score is just part of the calculation.
just one part of calculation.


```{r echo=FALSE, Multivariate_Plots2}
#subset NA values for ProsperRAting
mydata1<-subset(mydata, !is.na(ProsperRating..numeric.))

p<-ggplot(aes(x=LoanOriginalAmount, y=BorrowerRate,
       color=factor((ProsperRating..numeric.))),
       data=mydata1,  na.rm = TRUE)+
  geom_jitter()+ stat_ellipse()+scale_color_brewer(type = 'seq', palette = 'YlGnBu')+
  theme_gray(base_size = 14)
p
```

Here we see that borrower rate higher for loans with "bad" ratings. Big loans
have low Borrower rate and high ratings(5-7), while small loans(up to $10000) have all 
range of rates and ratings. Grey dots is missing data for Prosper Rating.
Let's see if there is dependency from Loan Status.

```{r echo=FALSE, Multivariate_Plots3}
ggplot(aes(x=LoanOriginalAmount, y=BorrowerRate,
       color=factor((ProsperRating..numeric.))),
       data=mydata1,  na.rm = TRUE)+
  geom_jitter()+scale_color_brewer(type = 'seq', palette = 'YlGnBu')+
  theme_gray(base_size = 14)+facet_wrap(~LoanStatus)
```

Rate, Rating and Amount don't depend on Loan Status. With amount it's
obvious, because it fixed. Probably, Borrower Rate and Rating is fixed also from the
moment of loan approval. In other practicies borrower's credit rating can go down for 
past due. 



# Multivariate Analysis

In this part of the investigation we confirm findings from Bivariate analysis 
and find a few new facts. For example, that Borrower Rate higher for loans with
"bad" ratings and low for "good ratings", but for loans up to $10000 it has all 
range of rates and ratings. This can testify that small amount of money can be 
given for different clients on different conditions, even with low Credit Ratings
considering other factors (monthly income, homeowner, etc). 
Also we found that low Prosper rating lead to bigger loss while Credit
Score Lower Range sometimes can be pretty high for such loans. 
This is one of the reasons, I think, that Prosper use their own rating.
And we found that there is no dependency between Loan Rate, Prosper Rating and 
Loan status, what means that they fixed and don't change when loan is past due.




# Final Plots and Summary


### Plot One

```{r echo=FALSE, Plot_One}
boxplot(LoanOriginalAmount~factor(ProsperRating..numeric.), data=mydata, 
        col = c("pink", "blue"), xlab = "ProsperRating",
         ylab = "Loan Amount, $",
         ylim = c(0, 25000) , main="Prosper Rating vs Loan Amount")

```


### Description One

This  plot shows how Prosper rating connected with Loan Amount.
Loans with high ammount of money has higher ratings, that's mean less risky.
High risk loans (HR Rating - 1) are loans for small amounts of money
(mostly up to $4,000) with some ouliers.

Also we have the same plots for Credit Score and Estimated Loss.


### Plot Two

```{r echo=FALSE, Plot_Two}
mydata1<-subset(mydata, !is.na(ProsperRating..numeric.))
  
ggplot(mydata1, aes(x=EstimatedLoss, y=CreditScoreRangeLower, 
                    fill=factor(ProsperRating..numeric.), colour=factor(ProsperRating..numeric.))) +
  geom_boxplot(alpha=0.6)+
  ggtitle("Distribution of Prosper Rating based on Credit Score and Estimated Loss")+
  labs(x = "EstimatedLoss(Rate)")
  
```

### Description Two
This plot I choose because it shows that Credit Score, that provided by credit 
rating agencies is not enough to prognose losses, that's why, probably, it was one 
of the factor why Prosper choose to use their own rating, which based on Credit Score.

### Plot Three

```{r echo=FALSE, Plot_Three}
mydata1<-subset(mydata, !is.na(ProsperRating..numeric.))

p<-ggplot(aes(x=LoanOriginalAmount, y=BorrowerRate,
       color=factor((ProsperRating..numeric.))),
       data=mydata1,  na.rm = TRUE)+
  geom_jitter()+ stat_ellipse()+scale_color_brewer(type = 'seq', palette = 'YlGnBu')+
  theme_gray(base_size = 14)+
       ggtitle("Distribution of ProsperRating by BorrowerRate and LoanAmount")+
       labs(x = "Loan Amount, $")
p
```

### Description Three
I choose this plot because it shows that big loans only with good ratings and good rates.
But as for smaller amounts rule still is, but it can vary sometimes and depend on other
factors.


# Reflection

Prosper loan data is pretty big data set with various parameters that represents
data from the end of 2005 to the 2014. One of the problems was dertermining which 
variables to analyze, how to find and deal with missing data and
not drifting too far off any one path. I started by understanding the individual
variables in the data set, and then I explored in more details Prosper Rating, that 
was created by this company, and parameteres that closely connected with it(based on 
correlation and rules of rating's calculation). 
Other problem for big data set is overplotting.
As I  make observations on plots I made conclusion that Prosper decide to use their
own rating because ratings from third companies are not enough to prognose losses.
It's also proved by historical data when  Prosper had to relaunch after negative return
and and implement stricter credit guidelines for borrowers in July 2009. 
Due to this fact my investigation limited by that fact that Prosper Rating is 
applicable for loans originated after July 2009.
There are a lot more things that we can investigate beased on this dataset.  
For example, we can compare some loan characteristic before and after new rules 
implementation. We could confirm that Prosper's loan default rate has been 
significantly reduced and the percentage of all loans that are 6+ months old,
and are 1+ month late, dropped to less than 4% (according to Wikipedia). 
It would be interesting to build a predictive model that quantifies if the person 
is going to default on their loan. We could use, for example, a logistic regression. 


